Analyzing portfolio effects in commercial catch data

### 2016-03-18 analyze_cfec_data.Rmd
update

---
title: "Analyzing portfolio effects in commercial catch data"
author: "NCEAS commercial catch portfolio working group"
date: "October 30, 2015"
output:
  pdf_document:
    fig_caption: yes
---

@@

```{r chunkSet, cache=FALSE, echo=FALSE,warning=FALSE,message=FALSE}
library(knitr)
library(dplyr)
library(ggplot2)
library(vegan)
library(mgcv)
opts_chunk$set(tidy=TRUE, message=FALSE)
```

@@

### Note: for any confidentiality concerns, the workspace loaded here has been slightly jittered. This will need to be replaced with the real data for the final model. 
```{r loadData}
load("cfec_jitter.Rdata")
# initially only use data 1985+
cfec = cfec[cfec$year>1984,]
```

The first data processing step is to aggregate some of the species. This is largely needed because of changes in reporting over time. For example, some of the rockfishes weren't reported to the same species level. For other species, there are unidentified groups, so there's additional aggregation required.

@@

# Lump flatfish (Shallow)
cfec$specn[cfec$specn==123] = 119 # rock
cfec$specn[cfec$specn==129] = 119 # starry flounder
cfec$specn[cfec$specn==127] = 119 # yellowfin
cfec$specn[cfec$specn==126] = 119 # butter
cfec$spec[cfec$specn==119] = "FLAT.shallow"

# Lump flatfish (deep)
cfec$specn[cfec$specn==124] = 118 # dover
cfec$specn[cfec$specn==134] = 118 # greenland turbot
cfec$spec[cfec$specn==118] = "FLAT.deep"

@@

# Lump king crabs
cfec$spec[cfec$spec%in%c("GKCR","BKCR","UKCR","RKCR")] = "KCRB" # king
# Lump tanner crabs
cfec$spec[cfec$spec%in%c("UTCR","BTCR")] = "TCRB" # tanner
# Lump shrimp - spot prawn, pink shrimp, 
cfec$spec[cfec$spec%in%c("SPRW","PSHP","USRM")] = "USRM" # shrimp

# drop records with no spp
cfec = cfec[-which(cfec$spec==""),]
```

@@

```{r}
# Filter out sub-districts and group things at larger areas
# Lump statistical areas by district -- groundfish
cfec$stat6 = as.character(as.numeric(cfec$stat6)) # as.numeric needed because of dumb integer
cfec$areaType = NA
cfec$stat6_char = NA
indx = which(cfec$stat6 > 99999 & cfec$stat6 < 999999)
cfec$stat6_char[indx] = nchar(cfec$stat6[indx])
cfec$areaType[indx] = "groundfish"
# This is just turning the last digit - sub region - into 0
cfec$stat6[indx] = paste(substr(cfec$stat6[indx],1,5), "0",sep="")
cfec$stat6[indx] = as.numeric(cfec$stat6[indx])
# Now do salmon / shellfish - This is just turning the last 2 digits - sub region - into 00
indx = which(cfec$stat6 > 9999 & cfec$stat6 < 100000)
cfec$stat6[indx] = paste(substr(cfec$stat6[indx],1,3), "00",sep="")
cfec$areaType[indx] = "salmonShellfish"

cfec$stat6 = as.numeric(cfec$stat6)
# Deal with the IPHC areas too. We can verify that nearly all spp caught in 3-digit areas are halibut,
#sort(table(cfec$spec[which(cfec$stat6 < 1000)]))
```

Next, we'll aggregate the stat6 codes into larger geographic regions. The focus here is on Prince William Sound, but for contrast we'll include Southeast Alaska, Kodiak W / E, and Cook Inlet.

@@

@@

Drop all species records that have less than 1000 total observations. This is discarding about 5000 of 20 million records. 
```{r discardSpec}
cfecClean = group_by(cfec, spec) %>%
  mutate(nspecies = n()) %>%
  filter(nspecies > 1000) %>%
  data.frame

# Also drop out unid groundfish, unid flatfish, smelts, grenadier, english sole -- all very rare
cfecClean = cfecClean[-which(cfecClean$spec%in%c("UFLT","UGRN","SMLT","GRDR","EGLS")),]

@@

Delete catch from fisheries that have less than 100 total records. This simplifies the dimension of the fishery codes a lot (from 378 to 234), removes about 3050 records, and doesn't affect the total earnings (these fisheries represent 0.0009 of total g_earn).  
```{r discardFishery}
cfecClean = group_by(cfecClean, p_fshy) %>%
  mutate(nfshy = n()) %>%
  filter(nfshy > 100) %>%
  data.frame
```

Subset the data to only include vessels that fished at any point in Prince William Sound.  
```{r useOnlyPWS}
cfecPWS = group_by(cfecClean, adfg) %>%
  mutate(in_PWS = any(area%in%"PWS")) %>%
  as.data.frame() %>%
  filter(in_PWS) %>%
  data.frame
```

Alan suggested looking at some coarse plots. First, we can show participation of vessels over time. Halibut fishing pre-1995 needed to get quota by catching halibut in 1988-1990.  
```{r,fig.cap="Time series of participation by year",fig.pos="placeHere"}
plot(min(cfecPWS$year):max(cfecPWS$year), specnumber(table(cfecPWS$year,cfecPWS$adfg)), xlab="",ylab="# Vessels", type="b")
```

We can also look at species by year

```{r,fig.cap="Time series of species caught by year",fig.pos="placeHere"}
# total species
par(mfrow = c(2,1), mgp = c(2,1,0), mai=c(0.5,0.5,0.3,0.1))
plot(min(cfecPWS$year):max(cfecPWS$year), specnumber(table(cfecPWS$year, cfecPWS$spec)), xlab="",ylab="# Species", type="b",main="Catch from vessels that ever fished in PWS")

plot(min(cfecPWS$year):max(cfecPWS$year), specnumber(table(cfecPWS$year[cfecPWS$area=="PWS"], cfecPWS$spec[cfecPWS$area=="PWS"])), xlab="",ylab="# Species", type="b",main="Catch only in PWS")
```

Look at species diversity using Vegan. How has diversity in # species fished changed over time?   
```{r veganSpecies,echo=FALSE, fig.cap="Time series of species diversity",fig.pos="placeHere"}
nSpecies.mean = NA # mean richness
nSpecies.cv = NA # cv richness
simp.mean = NA # mean simpson
simp.cv = NA # cv simpson
shannon.mean = NA # mean shannon
shannon.cv = NA # cv shannon
for(yr in min(cfecPWS$year):max(cfecPWS$year)) {
  nSpecies.mean[yr+1-min(cfecPWS$year)] = mean(specnumber(table(cfecPWS$adfg[cfecPWS$year==yr], cfecPWS$spec[cfecPWS$year==yr])))
   nSpecies.cv[yr+1-min(cfecPWS$year)] = sd(specnumber(table(cfecPWS$adfg[cfecPWS$year==yr], cfecPWS$spec[cfecPWS$year==yr])))/nSpecies.mean[yr+1-min(cfecPWS$year)]
  simp.mean[yr+1-min(cfecPWS$year)] = mean(diversity(table(cfecPWS$adfg[cfecPWS$year==yr], cfecPWS$spec[cfecPWS$year==yr]),index="invsimpson"))
  simp.cv[yr+1-min(cfecPWS$year)] = sd(diversity(table(cfecPWS$adfg[cfecPWS$year==yr], cfecPWS$spec[cfecPWS$year==yr]),index="invsimpson"))/simp.mean[yr+1-min(cfecPWS$year)]
  shannon.mean[yr+1-min(cfecPWS$year)] = mean(diversity(table(cfecPWS$adfg[cfecPWS$year==yr], cfecPWS$spec[cfecPWS$year==yr])))
  shannon.cv[yr+1-min(cfecPWS$year)] = sd(diversity(table(cfecPWS$adfg[cfecPWS$year==yr], cfecPWS$spec[cfecPWS$year==yr])))/simp.mean[yr+1-min(cfecPWS$year)]  
}

@@

We can do similar plots for the number of gears used / vessel  
```{r veganGears,echo=FALSE, fig.cap="Time series of diversity of number of fishing gears",fig.pos="placeHere"}
nGears.mean = NA # mean richness
nGears.cv = NA # cv richness
simp.mean = NA # mean simpson
simp.cv = NA # cv simpson
shannon.mean = NA # mean shannon
shannon.cv = NA # cv shannon
for(yr in min(cfecPWS$year):max(cfecPWS$year)) {
  nGears.mean[yr+1-min(cfecPWS$year)] = mean(specnumber(table(cfecPWS$adfg[cfecPWS$year==yr], cfecPWS$gearn[cfecPWS$year==yr])))
   nGears.cv[yr+1-min(cfecPWS$year)] = sd(specnumber(table(cfecPWS$adfg[cfecPWS$year==yr], cfecPWS$gearn[cfecPWS$year==yr])))/nGears.mean[yr+1-min(cfecPWS$year)]
  simp.mean[yr+1-min(cfecPWS$year)] = mean(diversity(table(cfecPWS$adfg[cfecPWS$year==yr], cfecPWS$gearn[cfecPWS$year==yr]),index="invsimpson"))
  simp.cv[yr+1-min(cfecPWS$year)] = sd(diversity(table(cfecPWS$adfg[cfecPWS$year==yr], cfecPWS$gearn[cfecPWS$year==yr]),index="invsimpson"))/simp.mean[yr+1-min(cfecPWS$year)]
  shannon.mean[yr+1-min(cfecPWS$year)] = mean(diversity(table(cfecPWS$adfg[cfecPWS$year==yr], cfecPWS$gearn[cfecPWS$year==yr])))
  shannon.cv[yr+1-min(cfecPWS$year)] = sd(diversity(table(cfecPWS$adfg[cfecPWS$year==yr], cfecPWS$gearn[cfecPWS$year==yr])))/simp.mean[yr+1-min(cfecPWS$year)]  
}

@@

## Metric of change
We can use absolute or relative earnings over time, by vessel. We adopted the relative measure, because we're comparing vessels of very different size classes. Sizes of individual vessels may change slowly over time. The earning metric is calculated as the first-differenced log earnings from time (t+1) to time (t), $log(earn_{t+1}) - log(earn_{t})$.

The key challenge here is that we need a metric of earnings that will allow us to compare different types of vessels (e.g. large, small) that don't overlap in time. For example, we need a metric that standardizes catches or earnings from a small vessel fishing 1985-1990 to that of a large vessel fishing 2000-2005. The relative change is universal across vessels, and can be interpreted as percent change (e.g. a value of -1 = exp(-1) = 0.367, or a ~ 63% decline in earnings between years).

@@

```{r mergeYears}
dummyDF = expand.grid(adfg = unique(totalPWS$adfg), year = min(totalPWS$year):max(totalPWS$year), stringsAsFactors = F)
# expand the data frame w/missing years
totalPWSExp = left_join(dummyDF, totalPWS) %>%
  arrange(adfg, year)
totalPWSExp = group_by(totalPWSExp, adfg) %>%
  mutate(difflog = c(NA, diff(log(totalearnings))), 
         diffspec = c(NA, diff(nspecies)),
         diffgear = c(NA, diff(ngear)),
         diffarea = c(NA, diff(narea)),
         difffishery = c(NA, diff(nfishery))) %>%
  filter(!is.na(difflog))
```

@@

```{r,fig.cap="Histogram of predictors (across all years and data)",fig.pos="placeHere"}
par(mfrow = c(2,2), mai=c(0.6,0.6,0.1,0.1))
#hist(log(totalPWSExp$totalearnings), main="Log of earnings",xlab="",col="grey")
hist(totalPWSExp$nspecies, main="Number of species", xlab="", col="grey")
hist(totalPWSExp$ngear, main="Number of gears", xlab="", col="grey")
hist(totalPWSExp$narea, main="Number of areas", xlab="", col="grey")
hist(totalPWSExp$nfishery, main="Number of fisheries", xlab="", col="grey")
```

We can also look at change in these predictors over time -- we can also look at the response, which is change in earnings from time t to time t+1.
```{r,fig.cap="Histogram of change in predictors (across all years and data)",fig.pos="placeHere"}
par(mfrow = c(3,2), mai=c(0.6,0.6,0.2,0.1))
hist(totalPWSExp$difflog, main="Log of change in earnings", xlab="", col="grey")
hist(totalPWSExp$diffspec, main="Change in species", xlab="", col="grey")
hist(totalPWSExp$diffgear, main="Change in gears", xlab="", col="grey")
hist(totalPWSExp$diffarea, main="Change in areas", xlab="", col="grey")
hist(totalPWSExp$difffishery, main="Change in fisheries", xlab="", col="grey")
```

## Plotting change in earnings through time

@@

```{r,fig.cap="Median and quartiles of change in earnings, by species, for individuals that don't change number of species fished through time",fig.pos="placeHere"}
group_by(totalPWSExp[totalPWSExp$diffspec==0,], year, nspecies) %>% 
  summarize(medDiffLog = median(difflog),  
  lower = quantile(difflog,probs = 0.25), 
  upper = quantile(difflog,probs = 0.75)) %>% 
ggplot(aes(year, medDiffLog)) + geom_line() + facet_wrap(~nspecies) +  geom_ribbon(aes(ymin=lower, ymax=upper), alpha=0.2)
```

```{r,fig.cap="Median and quartiles of change in earnings, by gears, for individuals that don't change number of gears fished through time",fig.pos="placeHere"}
group_by(totalPWSExp[totalPWSExp$diffgear==0,], year, ngear) %>% 
  summarize(medDiffLog = median(difflog), 
  lower = quantile(difflog,probs = 0.25), 
  upper = quantile(difflog,probs = 0.75)) %>% 
ggplot(aes(year, medDiffLog)) + geom_line() + facet_wrap(~ngear) +  geom_ribbon(aes(ymin=lower, ymax=upper), alpha=0.2)
```

```{r,fig.cap="Median and quartiles of change in earnings, by areas, for individuals that don't change number of areas fished through time",fig.pos="placeHere"}
group_by(totalPWSExp[totalPWSExp$diffarea==0,], year, narea) %>% 
  summarize(medDiffLog = median(difflog), 
  lower = quantile(difflog,probs = 0.25), 
  upper = quantile(difflog,probs = 0.75)) %>% 
ggplot(aes(year, medDiffLog)) + geom_line() + facet_wrap(~narea) +  geom_ribbon(aes(ymin=lower, ymax=upper), alpha=0.2)
```

>>>>

---
title: "Analyzing portfolio effects in commercial catch data"
author: "NCEAS commercial catch portfolio working group"
date: "February 22, 2016"
output:
  pdf_document:
    fig_caption: yes
  html_document: default
---

@@

```{r chunkSet, cache=FALSE, echo=FALSE,warning=FALSE,message=FALSE}
library(MASS)
library(knitr)
library(dplyr)
library(ggplot2)
library(vegan)
library(mgcv)
library(lme4)
library(scales)
library(fields)
library(reshape)
library(reshape2)
library(cluster)
opts_chunk$set(tidy=TRUE, message=FALSE)
```

@@

### Note: for any confidentiality concerns, the workspace loaded here has been slightly jittered. This will need to be replaced with the real data for the final model. 
```{r loadData}
load("cfec.Rdata")
# initially only use data 1985+
cfec = cfec[cfec$year>1984,]
```

The first data processing step is to aggregate some of the species. This is largely needed because of changes in reporting over time (before 1985 for example, lots of the catch was from foreign fleets & not reported to species level -- for example, some of the rockfishes). For other species, there are unidentified groups, so there's additional aggregation required.

@@

# Lump flatfish (Shallow)
cfec$specn[cfec$specn==123] = 119 # rock
cfec$specn[cfec$specn==129] = 119 # starry flounder
cfec$specn[cfec$specn==127] = 119 # yellowfin
cfec$specn[cfec$specn==126] = 119 # butter
cfec$spec[cfec$specn==119] = "FLAT.shallow"
# 5, 6, 7, 15, 26, 47, 61, 
# Lump flatfish (deep)
cfec$specn[cfec$specn==124] = 118 # dover
cfec$specn[cfec$specn==134] = 118 # greenland turbot
cfec$spec[cfec$specn==118] = "FLAT.deep"

@@

# Lump king crabs
cfec$spec[cfec$spec%in%c("GKCR","BKCR","UKCR","RKCR")] = "KCRB" # king
# Lump tanner crabs
cfec$spec[cfec$spec%in%c("UTCR","BTCR")] = "TCRB" # tanner
# Lump shrimp - spot prawn, pink shrimp, 
cfec$spec[cfec$spec%in%c("SPRW","PSHP","USRM")] = "USRM" # shrimp
# Correct REX and "REX "
cfec$spec[cfec$spec%in%c("REX ")] = "REX"
# drop records with no spp
cfec = cfec[-which(cfec$spec==""),]
```

Next, we have to clean up a few of the gears. Drop a few gears that are sporadically used, and combine longline gears.

```{r}
# 8 = fishwheel, used at start / end
# 13 = dipnet, only last 2 years
# 99 = other, stopped reporting
cfec = cfec[-which(cfec$gearn%in%c(8, 13, 99)),]
# Merge 6 and 61 into longline
cfec$gearn[which(cfec$gearn==61)] = 6
```

@@

```{r}
# Filter out sub-districts and group things at larger areas
# Lump statistical areas by district -- groundfish
cfec$stat6 = as.numeric(cfec$stat6) # as.numeric needed because of integer
cfec$areaType = NA
cfec$stat6_char = NA
indx = which(cfec$stat6 > 99999 & cfec$stat6 < 999999)
cfec$stat6_char[indx] = nchar(as.character(cfec$stat6[indx]))
cfec$areaType[indx] = "groundfish"
# This is just turning the last digit - sub region - into 0
cfec$stat6[indx] = as.numeric(paste(substr(as.character(cfec$stat6[indx]),1,5), "0",sep=""))
# Now do salmon / shellfish - This is just turning the last 2 digits - sub region - into 00
indx = which(cfec$stat6 > 9999 & cfec$stat6 < 100000)
cfec$stat6[indx] = as.numeric(paste(substr(as.character(cfec$stat6[indx]), 1, 3), "00", sep=""))
cfec$areaType[indx] = "salmonShellfish"

# Deal with the IPHC areas too. We can verify that nearly all spp caught in 3-digit areas are halibut,
#sort(table(cfec$spec[which(cfec$stat6 < 1000)]))
```

Next, we'll aggregate the stat6 codes into larger geographic regions. The focus here is on Prince William Sound, but for contrast we'll include Southeast Alaska, Kodiak W / E, Cook Inlet, and Bristol Bay.

@@

Another alternative approach to grouping would be to perform a cluster analysis on vessels / areas. Here we'll make a table of vessels (rows) and areas (column) then use 'clara' to construct 8 groups. This is somewhat arbitrary, but reflects roughly the number constructed above.

```{r}
T = table(cfec$adfg, cfec$stat6)
CC = clara(T, k = 8)
cluster_vesselArea = t(CC$medoids)
```

Next we can sort out the mutually exclusive areas in the medoids to identify how the areas are being clustered.  
```{r}
cluster_vesselArea = ceiling(cluster_vesselArea / 1.0e10)
cluster_vesselArea = cluster_vesselArea[-which(apply(cluster_vesselArea,1,sum)!=1),]
```

For example, the 6th medoid is clearly pulling out vessels in Southeast, 
```{r}
names(which(cluster_vesselArea[,6]==1))
```

Similarly, medoid 1 is PWS, medoid 2 is Cook Inlet, medoid 4 is Bristol Bay, medoid 5 is ~ 700 records of halibut (from a southern area not in GoA), medoid 6 is Southeast, medoid 7 is Kodiak, medoids 3 and 8 are Bering. So this seems to agree with our clustering, maybe with the exception of not splitting Kodiak into W/E.

We can look at how the species landed varies by area (table in 1000s), 
```{r}
kable(round(table(cfec$spec, cfec$area)/1000, 2))
```

We can filter out rare species, and turn this table into catch composition (by occurence, not weight or earnings) to highlight how the areas are different,

```{r}
specArea = table(cfec$spec, cfec$area)
# normalize by area
for(i in 1:dim(specArea)[2]) {
specArea[,i] = specArea[,i]/sum(specArea[,i])
}
kable(round(specArea[-which(apply(specArea,1,max) < 0.01),], 2))
```

@@

Drop all species records that have less than 1000 total observations. This is discarding about 5000 of 20 million records. 
```{r discardSpec}
cfecClean = group_by(cfec, spec) %>%
  mutate(nspecies = n()) %>%
  filter(nspecies > 1000) %>%
  data.frame
# Also drop out unid groundfish, unid flatfish, smelts, grenadier, english sole -- all very rare
cfecClean = cfecClean[-which(cfecClean$spec%in%c("UFLT","UGRN","SMLT","GRDR","EGLS")),]
```

Delete catch from fisheries that have less than 100 total records. This simplifies the dimension of the fishery codes a lot (from 378 to 234), removes about 3050 records, and doesn't affect the total revenue (these fisheries represent 0.0009 of total g_earn).  
```{r discardFishery}
cfecClean = group_by(cfecClean, p_fshy) %>%
  mutate(nfshy = n()) %>%
  filter(nfshy > 100) %>%
  data.frame

# Write the SSL related sites to a workspace
sslStat6 = c(355630,
365730, 375830, 345530, 335430, 335430, 335430,
335430, 335430, 365800, 355700, 345530, 355630,
355700, 335700, 335700, 355700, 445930, 495930, 
476000, 485930, 495930, 325600, 495930, 476000,
475930, 555530, 565600, 525700, 525700, 525830,
525700, 515800, 515830, 535630, 365800, 575630,
535630, 565630, 525830, 595500, 595430, 595500,
595430, 635430, 635430, 605430, 615430, 635504,
585500)

# calculate areas in circle around square of interest
sslStat6.square = c(sslStat6, sslStat6 +- 9970, sslStat6 +- 10000, 
sslStat6 +- 10070, sslStat6 +- 70, sslStat6 +- 9930, sslStat6 +- 10000, 
sslStat6 +- 10030, sslStat6 +- 30)
sslStat6.label = rep(sslStat6,9)

cfecSSL = cfecClean[cfecClean$stat6%in%sslStat6.square,]
cfecSSL$label = sslStat6.label[match(cfecSSL$stat6, sslStat6.square)]
save(cfecSSL, file="cfecSSL.Rdata")

@@

Tabulate the gears and species remaining.  
```{r tableOfGears}
m = as.matrix(sort(table(cfecClean$gearn)), ncol=1)
# Normalize percent of all records
m = cbind(m, round(100*m[,1]/sum(m[,1]),2))
colnames(m) = c("Records","Percent")
kable(m)
```

This shows that 79% of the records in the database are salmon!   
```{r tableOfSpec}
m = as.matrix(sort(table(cfecClean$spec)), ncol=1)
# Normalize percent of all records
m = cbind(m, round(100*m[,1]/sum(m[,1]),2))
colnames(m) = c("Records","Percent")
kable(m)
```

Subset the data to only include vessels that ever fished in Prince William Sound. This data frame becomes 'cfecPWS'.

```{r useOnlyPWS}
cfecPWS = group_by(cfecClean, adfg) %>%
  mutate(in_PWS = any(area%in%"PWS")) %>%
  as.data.frame() %>%
  filter(in_PWS) %>%
  data.frame
```

Alan suggested looking at some coarse plots. First, we can show participation of vessels over time. Halibut fishing pre-1995 needed to get quota by catching halibut in 1988-1990.  
```{r,fig.cap="Time series of participation by year",fig.pos="placeHere"}
plot(min(cfecPWS$year):max(cfecPWS$year), specnumber(table(cfecPWS$year,cfecPWS$adfg)), xlab="",ylab="# Vessels", type="b", main = "Participation in PWS region")
```

We can also look at species landed by year. This is not terribly informative -- for example the total number of species landed in the dataset has ranged from 38-41. An alternative would be to take the average (below).

```{r,fig.cap="Time series of species caught by year",fig.pos="placeHere"}
# total species
par(mfrow = c(3,1), mgp = c(2,1,0), mai=c(0.5,0.5,0.3,0.1))
plot(min(cfecPWS$year) : max(cfecPWS$year) , 
     specnumber( table (cfecPWS$year, cfecPWS$spec)), xlab="", 
     ylab="# Species", type="b", 
     main="Catch from vessels that ever fished in PWS")

plot(min(cfecPWS$year) : max(cfecPWS$year), 
     specnumber(table(cfecPWS$year[cfecPWS$area=="PWS"], 
                      cfecPWS$spec[cfecPWS$area=="PWS"])), 
     xlab="",ylab="# Species", type="b",main="Catch only in PWS")

plot(min(cfecClean$year) : max(cfecClean$year), 
     specnumber(table(cfecClean$year,  cfecClean$spec)), 
     xlab="",ylab="# Species", type="b", 
     main="All vessels in dataset")
```

Now we can calculate the average species landed / year, where the average is calculated as the total number of species landed divided by the total number of vessels fishing.

```{r,fig.cap="Time series of avg species caught by year per vessel",fig.pos="placeHere"}
# total species
par(mfrow = c(3,1), mgp = c(2,1,0), mai=c(0.5,0.5,0.3,0.1))
plot(min(cfecPWS$year):max(cfecPWS$year), 
     specnumber(table(cfecPWS$year, cfecPWS$spec)) / 
       specnumber(table(cfecPWS$year,  cfecPWS$adfg)), 
     xlab="", ylab="Avg # Species / Ves", 
     type="b", main="Catch from vessels that ever fished in PWS")

plot(min(cfecPWS$year):max(cfecPWS$year), 
     specnumber(table(cfecPWS$year[cfecPWS$area=="PWS"], 
     cfecPWS$spec[cfecPWS$area=="PWS"])) / 
       specnumber(table(cfecPWS$year[cfecPWS$area=="PWS"], 
     cfecPWS$adfg[cfecPWS$area=="PWS"])), 
     xlab="",ylab="Avg # Species / Ves", 
     type="b",main="Catch only in PWS")

plot(min(cfecClean$year):max(cfecClean$year), 
     specnumber(table(cfecClean$year,  cfecClean$spec)) / 
       specnumber(table(cfecClean$year,  cfecClean$adfg)), 
     xlab="",ylab="Avg # Species / Ves",
     type="b", main="All vessels in dataset")
```

Another way to calculate the average is total # species landed divided by the total number of fishing records per year.

```{r,fig.cap="Time series of avg species caught by year per vessel",fig.pos="placeHere"}
# total species
par(mfrow = c(3,1), mgp = c(2,1,0), mai=c(0.5,0.5,0.3,0.1))
plot(min(cfecPWS$year):max(cfecPWS$year), 
     specnumber(table(cfecPWS$year, cfecPWS$spec)) / 
       as.numeric(table(cfecPWS$year)), xlab="", 
     ylab = "Avg # Species / Haul", type = "b",  
     main = "Catch from vessels that ever fished in PWS")

plot(min(cfecPWS$year):max(cfecPWS$year), 
     specnumber(table(cfecPWS$year[cfecPWS$area=="PWS"], 
    cfecPWS$spec[cfecPWS$area=="PWS"])) / 
      as.numeric(table(cfecPWS$year[cfecPWS$area=="PWS"])), 
    xlab = "", ylab = "Avg # Species / Haul", 
    type = "b",main = "Catch only in PWS")

plot(min(cfecClean$year):max(cfecClean$year), 
     specnumber(table(cfecClean$year,  cfecClean$spec)) / 
       as.numeric(table(cfecClean$year)), xlab = "", 
     ylab = "Avg # Species / Haul", 
     type = "b", main = "All vessels in dataset")
```

Look at species diversity using Vegan. How has diversity in # species fished changed over time?   
```{r veganSpecies,echo=FALSE, fig.cap="Time series of species diversity (for vessels that have ever fished in Prince William Sound)",fig.pos="placeHere"}
nSpecies.mean = NA # mean richness
nSpecies.cv = NA # cv richness
simp.mean = NA # mean simpson
simp.cv = NA # cv simpson
shannon.mean = NA # mean shannon
shannon.cv = NA # cv shannon
for(yr in min(cfecPWS$year):max(cfecPWS$year)) {
  nSpecies.mean[yr+1-min(cfecPWS$year)] = mean(specnumber(table(cfecPWS$adfg[cfecPWS$year==yr], cfecPWS$spec[cfecPWS$year==yr])))
   nSpecies.cv[yr+1-min(cfecPWS$year)] = sd(specnumber(table(cfecPWS$adfg[cfecPWS$year==yr], cfecPWS$spec[cfecPWS$year==yr])))/nSpecies.mean[yr+1-min(cfecPWS$year)]
  simp.mean[yr+1-min(cfecPWS$year)] = mean(diversity(table(cfecPWS$adfg[cfecPWS$year==yr], cfecPWS$spec[cfecPWS$year==yr]),index="invsimpson"))
  simp.cv[yr+1-min(cfecPWS$year)] = sd(diversity(table(cfecPWS$adfg[cfecPWS$year==yr], cfecPWS$spec[cfecPWS$year==yr]),index="invsimpson"))/simp.mean[yr+1-min(cfecPWS$year)]
  shannon.mean[yr+1-min(cfecPWS$year)] = mean(diversity(table(cfecPWS$adfg[cfecPWS$year==yr], cfecPWS$spec[cfecPWS$year==yr])))
  shannon.cv[yr+1-min(cfecPWS$year)] = sd(diversity(table(cfecPWS$adfg[cfecPWS$year==yr], cfecPWS$spec[cfecPWS$year==yr])))/simp.mean[yr+1-min(cfecPWS$year)]  
}

@@

We can do similar plots for the number of gears used / vessel  
```{r veganGears,echo=FALSE, fig.cap="Time series of diversity of number of fishing gears (for vessels that have ever fished in Prince William Sound)",fig.pos="placeHere"}
nGears.mean = NA # mean richness
nGears.cv = NA # cv richness
simp.mean = NA # mean simpson
simp.cv = NA # cv simpson
shannon.mean = NA # mean shannon
shannon.cv = NA # cv shannon
for(yr in min(cfecPWS$year):max(cfecPWS$year)) {
  nGears.mean[yr+1-min(cfecPWS$year)] = mean(specnumber(table(cfecPWS$adfg[cfecPWS$year==yr], cfecPWS$gearn[cfecPWS$year==yr])))
   nGears.cv[yr+1-min(cfecPWS$year)] = sd(specnumber(table(cfecPWS$adfg[cfecPWS$year==yr], cfecPWS$gearn[cfecPWS$year==yr])))/nGears.mean[yr+1-min(cfecPWS$year)]
  simp.mean[yr+1-min(cfecPWS$year)] = mean(diversity(table(cfecPWS$adfg[cfecPWS$year==yr], cfecPWS$gearn[cfecPWS$year==yr]),index="invsimpson"))
  simp.cv[yr+1-min(cfecPWS$year)] = sd(diversity(table(cfecPWS$adfg[cfecPWS$year==yr], cfecPWS$gearn[cfecPWS$year==yr]),index="invsimpson"))/simp.mean[yr+1-min(cfecPWS$year)]
  shannon.mean[yr+1-min(cfecPWS$year)] = mean(diversity(table(cfecPWS$adfg[cfecPWS$year==yr], cfecPWS$gearn[cfecPWS$year==yr])))
  shannon.cv[yr+1-min(cfecPWS$year)] = sd(diversity(table(cfecPWS$adfg[cfecPWS$year==yr], cfecPWS$gearn[cfecPWS$year==yr])))/simp.mean[yr+1-min(cfecPWS$year)]  
}

@@

## Metric of change
We can use absolute or relative revenue over time, by vessel. We adopted the relative measure, because we're comparing vessels of very different size classes. Sizes of individual vessels may change slowly over time. The revenue metric is calculated as the first-differenced log revenue from time (t+1) to time (t), $log(earn_{t+1}) - log(earn_{t})$.

The key challenge here is that we need a metric of revenue that will allow us to compare different types of vessels (e.g. large, small) that don't overlap in time. For example, we need a metric that standardizes catches or revenue from a small vessel fishing 1985-1990 to that of a large vessel fishing 2000-2005. The relative change is universal across vessels, and can be interpreted as percent change (e.g. a value of -1 = exp(-1) = 0.367, or a ~ 63% decline in revenue between years).

@@

```{r mergeYears}
dummyDF = expand.grid(adfg = unique(totalPWS$adfg), 
  year = min(totalPWS$year) : max(totalPWS$year), 
  stringsAsFactors = F)
# expand the data frame w/missing years
totalPWSExp = left_join(dummyDF, totalPWS) %>%
  arrange(adfg, year)
totalPWSExp = group_by(totalPWSExp, adfg) %>%
  mutate(difflog = c(NA, diff(log(totalearnings))), 
         diffspec = c(NA, diff(nspecies)),
         diffgear = c(NA, diff(ngear)),
         diffarea = c(NA, diff(narea)),
         difffishery = c(NA, diff(nfishery))) %>%
  filter(!is.na(difflog))
```

@@

```{r,fig.cap="Histogram of predictors (across all years and data)",fig.pos="placeHere"}
par(mfrow = c(2,2), mai=c(0.6,0.6,0.1,0.1))
#hist(log(totalPWSExp$totalearnings), main="Log of revenue",xlab="",col="grey")
hist(totalPWSExp$nspecies, main="Number of species", xlab="", col="grey")
hist(totalPWSExp$ngear, main="Number of gears", xlab="", col="grey")
hist(totalPWSExp$narea, main="Number of areas", xlab="", col="grey")
hist(totalPWSExp$nfishery, main="Number of fisheries", xlab="", col="grey")
```

We can also look at change in these predictors over time -- we can also look at the response, which is change in revenue from time t to time t+1.
```{r,fig.cap="Histogram of change in predictors (across all years and data)",fig.pos="placeHere"}
par(mfrow = c(3,2), mai=c(0.6,0.6,0.2,0.1))
hist(totalPWSExp$difflog, main="Log of change in revenue", xlab="", col="grey")
hist(totalPWSExp$diffspec, main="Change in species", xlab="", col="grey")
hist(totalPWSExp$diffgear, main="Change in gears", xlab="", col="grey")
hist(totalPWSExp$diffarea, main="Change in areas", xlab="", col="grey")
hist(totalPWSExp$difffishery, main="Change in fisheries", xlab="", col="grey")
```

## Plotting change in revenue through time

@@

```{r,fig.cap="Median and quartiles of change in revenue, by species, for individuals that don't change number of species fished through time",fig.pos="placeHere"}
group_by(totalPWSExp[totalPWSExp$diffspec==0, ], year, nspecies) %>%  
  summarize(medDiffLog = median(difflog),  
  lower = quantile(difflog,probs = 0.25),  
  upper = quantile(difflog,probs = 0.75)) %>%  
ggplot(aes(year, medDiffLog)) + geom_line() + facet_wrap(~nspecies) +  geom_ribbon(aes(ymin=lower, ymax=upper), alpha=0.2)
```

```{r,fig.cap="Median and quartiles of change in revenue, by gears, for individuals that don't change number of gears fished through time",fig.pos="placeHere"}
group_by(totalPWSExp[totalPWSExp$diffgear==0, ], year, ngear) %>%  
  summarize(medDiffLog = median(difflog),  
  lower = quantile(difflog,probs = 0.25),  
  upper = quantile(difflog,probs = 0.75)) %>%  
ggplot(aes(year, medDiffLog)) + geom_line() + facet_wrap(~ngear) +  geom_ribbon(aes(ymin=lower, ymax=upper), alpha=0.2)
```

```{r,fig.cap="Median and quartiles of change in revenue, by areas, for individuals that don't change number of areas fished through time",fig.pos="placeHere"}
group_by(totalPWSExp[totalPWSExp$diffarea==0, ], year, narea) %>%  
  summarize(medDiffLog = median(difflog),  
  lower = quantile(difflog,probs = 0.25),  
  upper = quantile(difflog,probs = 0.75)) %>%  
ggplot(aes(year, medDiffLog)) + geom_line() + facet_wrap(~narea) +  geom_ribbon(aes(ymin=lower, ymax=upper), alpha=0.2)
```

```{r,fig.cap="Median and quartiles of change in revenue, by areas, for individuals that don't change number of fisheries fished through time",fig.pos="placeHere"}
group_by(totalPWSExp[totalPWSExp$difffishery==0, ], year, nfishery) %>%  
  summarize(medDiffLog = median(difflog),  
  lower = quantile(difflog,probs = 0.25),  
  upper = quantile(difflog,probs = 0.75)) %>%  
ggplot(aes(year, medDiffLog)) + geom_line() + facet_wrap(~nfishery) +   geom_ribbon(aes(ymin=lower, ymax=upper), alpha=0.2)
```
